---
output:
  pdf_document: default
urlcolor: blue
header-includes:    
  - \usepackage{lastpage}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[CO, CE]{Xiyuan Gu}
  - \fancyfoot[CO, CE]{\thepage \ of \pageref{LastPage}}
---

```{r setup, message = FALSE, echo=FALSE}
# Students: You probably shouldn't change any of the code in this chunk.

# These are the packages you will need for this activity
packages_needed <- c("tidyverse", "googledrive", "readxl", "janitor", 
                     "lubridate", "opendatatoronto", "ggthemes")

package.check <- lapply(
  packages_needed,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
    }
  }
)

# Credit: package.check based on a helpful post from Vikram Baliga https://vbaliga.github.io/verify-that-r-packages-are-installed-and-loaded/

# Load tidyverse
library(tidyverse)
library(readxl)
library(janitor)
library(ggthemes)

# Set so that long lines in R will be wrapped:
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), echo = TRUE)
```


```{r load_data,message=FALSE, echo=FALSE}
lifetime_earn <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/lifetime_earn.csv')
student_debt <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/student_debt.csv')
retirement <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/retirement.csv')
home_owner <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/home_owner.csv')
race_wealth <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/race_wealth.csv')
income_time <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_time.csv')
income_limits <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_limits.csv')
income_aggregate <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_aggregate.csv')
income_distribution <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_distribution.csv')
income_mean <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-09/income_mean.csv')

```

Code last run `r Sys.Date()`.  
Data: `r date('2021-02-09')`.   

# TidyTuesday: Income and Wealth Over Time

## About the data
The income and wealth data is from the Urban Institute and the US Census. The Urban Institute provides data on wealth by race and the US Census focuses on historical income.  The dataset in this Tidytuesday project contains `lifetime_earn`, `student_debt`, `retirement`, `home_owner`, `race_wealth`, `income_time`, `income_limits`, `income_aggregate`, `income_distribution` and `income_mean`of each race yearly. The illustration could be stacked bar plots, boxplots, line graphs for summary or comparisons.
What I am interested in visualizing from this dataset is to see how the wealth and income of minority groups such as black people and Asians distribute, compared with the White.

## Visualizations for Income

### Data Wrangling
I filtered the races into `All Races`, `Black Alone`, `White Alone`, `Hispanic (Any Race)` and `Asian Alone` for my comparison. Since `Asian Alone`  started to record in 1988, I visualized the distribution starting the year 1987. The filtered dataset is stored in `income_data`.

```{r}
asian<-income_distribution%>%filter(race=="Asian Alone") #from 1988 to 2019
income_data<-income_distribution%>%
  filter(race %in% c('All Races',"Black Alone", "White Alone", "Hispanic (Any Race)","Asian Alone")) %>%
  filter(year >= 1988)
na<-income_data%>%
  filter(is.na(income_mean))#no NAs in income_mean
na<-income_data%>%
  filter(is.na(income_distribution))#no NAs in income_distribution

```

### Median Income over time by Race
From Graph 1, we can see that the Asian (alone) ethnic group has the highest median of income of all time and Asian (alone) and White are the two groups above the median for all races.
Whereas Hispanic (all race) and Black (alone) are above the median of all races and Black (alone) has the lowest median of income of all time from 1988 to 2019.

```{r}
library(ghibli)
library(grid)
income_data$race <- factor(income_data$race , levels = c("Asian Alone", "White Alone",'Hispanic (Any Race)', 'Black Alone', "All Races"))
plot1<-income_data%>%ggplot(aes(x = year, y = income_median, group = race,color=race)) +geom_line()+
  scale_y_continuous(labels = scales::comma)+
  geom_text(data = filter(income_data, year == "2015"),
              aes(label = race),
              hjust = 0, nudge_x = 0.2) +
  labs(title = 'Graph 1: Median Income by Race, 1988-2019',
       x='Year',
       y='Income Median ($)',
       caption = str_c("Created by: Xiyuan Gu for TidyTuesday\nSource: the Urban Institute and the US Census\nData as of ", date('2021-02-09')))+
  theme_minimal()+ theme(legend.position =  'none')+
  scale_x_continuous(breaks = seq(1988, 2021, 3), limits = c(1988, 2022)) +
  scale_colour_ghibli_d("KikiMedium", direction = -1)
plot1
ggsave("plot1.png", plot = plot1)
```


### Income Distribution vs Income Bracket over time for `Black Alone`, `White Alone`, `Hispanic (Any Race)` and `Asian Alone`.
This interactive plot, Graph 2, shows the income distribution (%) of each ethnicity group in all income bracket from 1988 to 2019. The size of the points increases as the year goes by. The `Black alone` group occupies the largest proportion in the lowest income bracket, `Under $15,000`, and the Black's proportion decreases as the income bracket goes up. The `Hispanic (any race)` group also follows this trend. Whereas most of the `Asian alone`  distributes in the bracket of `$50,000 to $74,999` or above. Notably, over recent years, the income distribution of `Asian alone` rises in the highest income bracket, `$200,000 and over`, so does `White alone`. The income proportion of `White alone` mainly concentrates on the middle and lower range of income bracket and only about 20% proportion earn above $150,000.

```{r}
library(plotly)
library(ghibli)
library(ggplot2)
income_data$income_bracket = factor(income_data$income_bracket, levels=c('$200,000 and over','$150,000 to $199,999','$100,000 to $149,999','$75,000 to $99,999','$50,000 to $74,999','$35,000 to $49,999','$25,000 to $34,999','$15,000 to $24,999','Under $15,000'))
income_data$race = factor(income_data$race, levels=c("Asian Alone","White Alone", "Hispanic (Any Race)","Black Alone"))

p<-income_data%>%
  filter(race %in% c("Black Alone", "White Alone", "Hispanic (Any Race)","Asian Alone"))%>%
  ggplot()+geom_point(aes(x=income_distribution,y=income_bracket,color=race,group=race,size=year),alpha=0.5, position = position_dodge(width = 0.75))+
  scale_colour_ghibli_d("KikiMedium", direction = -1)+
  labs(title = 'Graph 2: Income Distribution by Race of each income bracket in US, 1988-2019',
       y='Income Bracket',
       x='Income Distribution (%)',
       caption = str_c("Created by: Xiyuan Gu for TidyTuesday\nSource: the Urban Institute and the US Census\nData as of ", date('2021-02-09')))+
  theme(legend.position = 'none',
        panel.grid.minor.x = element_blank())+
  theme_minimal()
plot2<-ggplotly(p)

```

```{r,echo=FALSE}
# save the widget at .html format
library(htmlwidgets)
saveWidget(plot2, file="/users/lucygu/downloads/plot2.html")
```

### line graph of income_time
This interactive plot, Graph 3, shows the family incomes with `upper-income status` (90th income percentile) have increased from `$95,000` in 1963 to `$180,000` in 2016, representing a twofold increase. For  `middle-income status` (50th income percentile) families, their incomes have slowly increased from `$47,000` to `$65,000`. However,  `low-income status` (10th income percentile) families remain familial incomes of around `$14,000`. Overall, the gaps between the upper and lower familial income have widened over time. 

```{r}
library(plotly)
fam_income<-income_time%>%
  group_by(year)%>%
  mutate(income_status=case_when(
    percentile=="10th"~"Low-income Status",
    percentile=="50th"~"Middle-income Status",
    percentile=="90th"~"Upper-income Status"))

fig <- fam_income%>%ggplot()+
  geom_line(aes(x = year, y = income_family,color=income_status))+
  geom_point(aes(x = year, y = income_family,color=income_status))+
  scale_colour_ghibli_d("KikiMedium", direction = -1)+
  labs(title = 'Graph 3: Family-level income by percentile from 1963 to 2016',
       y='Familial income ($)',
       caption = str_c("Created by: Xiyuan Gu for TidyTuesday\nSource: the Urban Institute and the US Census\nData as of ", date('2021-02-09')))+theme_minimal()
plot3<-ggplotly(fig)
plot3 <- plot3%>%
  layout(hovermode = "x unified")
         
plot3
```


```{r,echo=FALSE}
# save the widget at .html format
library(htmlwidgets)
saveWidget(plot3, file="/users/lucygu/downloads/plot3.html")
```

## Visualizations for Wealth

## Visualization for Wealth By gender and race
Graph 4 shows that the average lifetime earnings vary between different races and genders. The largest gender disparity of average lifetime earning is between White women and White men for about $1,000,000 and the smallest is between Black women and men. However, regarding races, both  White women's and men's average lifetime earnings are higher compared with Hispanic's and Black's.

```{r}
library(ggplot2)
library(ggthemes)
library(dplyr)
library(ghibli)

brks <- seq(-3000000, 3000000, 1000000)
lbls = paste0(as.character(c(seq(3, 0, -1), seq(1, 3, 1))), "m")

plot4<-ggplot(data=lifetime_earn,aes(y=lifetime_earn,x=race,fill=gender)) + 
  geom_bar(data=subset(lifetime_earn,gender=="Women"),stat = "identity", width = .6) + 
  geom_bar(data=subset(lifetime_earn,gender=="Men"),
           stat = "identity", width = .6,aes(y=lifetime_earn*(-1))) + 
  scale_y_continuous(breaks=brks,labels=lbls) + 
  scale_fill_ghibli_d("LaputaMedium", direction = -1)+
  labs(title="Graph 4: Average lifetime earning by gender and race", 
       x='Race',
       y='Average lifetime earning',
       caption = str_c("Created by: Xiyuan Gu for TidyTuesday\nSource: the Urban Institute and the US Census\nData as of ", date('2021-02-09')))+
  
  guides(fill=guide_legend(title = NULL))+
  theme_minimal()+coord_flip()

plot4
ggsave("plot4.png", plot = plot4)
```

### Visualization of family wealth between White and Black ethnic groups

In 1983, white families had `$92,045` in median more wealth normalized to 2016 than black families. By 2016, the disparity in family wealth grew to `$153,591`. Notably, the biggest disparity between white and black families is `$174,304` in 2007. When the global recession continues, the median wealth in white families dropped dramatically by `$50,000` in the next three years while in black families only dropped a little.

```{r}
wealth_wb<-race_wealth%>%
  filter(!is.na(wealth_family))%>%
  filter(type=="Median")%>%
  filter(race %in% c("White","Black"))%>%
  filter(year>=1980)
#for ribbon
ribbon<-wealth_wb%>%
  pivot_wider(names_from = race, values_from = wealth_family)%>%
  mutate(disparity=White-Black)
# for label disparity
text_1983 <- ribbon %>%
  mutate(position = 60000) %>%
  filter(year == 1983)
text_2007 <- ribbon %>%
  mutate(position = 100000) %>%
  filter(year == 2007)
text_2016 <- ribbon %>%
  mutate(position = 90000) %>%
  filter(year == 2016)

p5<-ggplot()+
  geom_line(data=wealth_wb,aes(x=year,y=wealth_family,color=race))+
  geom_ribbon(data=ribbon,aes(x=year,ymin = Black, ymax = White), 
              fill = ghibli_palette("SpiritedLight",direction = -1)[1], 
              alpha = 0.5)+ 
  
  #annotate labels positions
  annotate("text", x = 2014.5, y = 175000, 
         label = "White Families", 
         color = ghibli_palette("KikiMedium", direction = -1)[2], 
         size = 3, hjust = 0, vjust = .5) +
   annotate("text", x = 2014.5, y = 12500, 
         label = "Black Families", 
         color = ghibli_palette("KikiMedium", direction = -1)[1], 
         size = 3, hjust = 0, vjust = .5) +
  # add segment lines (disparity markers)
  geom_segment(data = ribbon, aes(x = year,xend = year,y = White,yend = Black),
               linetype = "dotted",
               color = ghibli_palette("PonyoLight",direction = -1)[5]) +
 
   # add label for 1983, 2007, 2016
  geom_label(data = text_2007,
             aes(x = year,y = position,
                 label =paste0("$", round(disparity,0))),
             color = "#f9fafc", 
             fill = ghibli_palette("SpiritedDark",direction = -1)[1], size = 2.5) +
    geom_label(data = text_1983,
             aes(x = year,y = position,
                 label =paste0("$", round(disparity,0))),
             color = "#f9fafc", 
             fill = ghibli_palette("SpiritedDark",direction = -1)[1], size = 2.5) +
    geom_label(data = text_2016,
             aes(x = year,y = position,
                 label =paste0("$", round(disparity,0))),
             color = "#f9fafc", 
             fill = ghibli_palette("SpiritedDark",direction = -1)[1], size = 2.5) +
  
  labs(title="Graph 5: Disparity in family wealth normalized to 2016, 1983-2016", 
       x='Year',
       y='Median Family Wealth',
       caption = str_c("Created by: Xiyuan Gu for TidyTuesday\nSource: the Urban Institute and the US Census\nData as of ", date('2021-02-09')))+
   theme_minimal()+ theme(legend.position =  'none')+
  scale_x_continuous(breaks = seq(1983, 2016, 3), limits = c(1982, 2017)) +
  scale_colour_ghibli_d("KikiMedium", direction = -1)
p5
ggsave("plot5.png", plot = p5)
```

